---
title: "Project1"
output: html_document
header-includes:
   - \usepackage{amsmath}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("main.R")
```
\newcommand{\vect}[1]{\boldsymbol{\mathbf{#1}}}
\newcommand{\matr}[1]{\boldsymbol{\mathbf{#1}}}
\newcommand{\det}{det}
\newcommand{\exp}{exp}
$$E(r(X)) = \mu_r = 0$$

$$Var(r(X)) = \sigma_r^2$$
 $$ Cor(r(X), r(X')) = \rho_r(\tau) $$
 
 $$\tau = \frac{| X - X'|}{10}$$
 
 $(r(x) | x\in D = [1, 50]\subset \mathbb{R}^1)$
 
 $\vect{X}$
```{r}
p1
p2
```

 # Problem 1b)
 Want to specify pdf of a guassian prior rf.

Let $\vect r \sim p(\vect r) = \phi_n(\vect r | \mu_r\vect 1_n, \sigma^2_r \matr \Sigma_r^\rho$
With $\matr \Sigma_r^p$ covariance matrix defined by the corrolation function given in problem 1a) for a given $\kappa$. To start off we select $\kappa = 0.5$.

This would give use the following pdf:
$$f_{\vect r}(\vect r | \mu_r, \sigma_r^2\matr \Sigma_r^\rho)= (2\pi)^{-\frac{n}{2}}\det(\matr\Sigma_r^\rho)^{-\frac{1}{2}}\exp\left(-\frac{1}{2}\left(\vect r - \mu_r\vect 1_n\right)^T\matr\Sigma_r^\rho\left(\vect r - \mu_r\vect 1_n\right)\right)$$